---
title: "Evaluating informative hypotheses using bain"
author: "Caspar van Lissa"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
categories: ["R"]
tags: ["R", "Bayes", "SEM"]
bibliography: [bibliography.bib]
link-citations: true
csl: apa.csl
---

I recently presented on *Evaluating informative hypotheses for parameters of a structural equation model using Bayes factors* at [EARA 2020](https://www.fpce.up.pt/eara2020/home.html). This blog post was released in tandem with the conference presentation, to provide a brief hands-on tutorial on how to conduct these analyses yourself.

The YouTube video below demonstrates how to evaluate informative hypotheses using `bain`. If you want to follow along, you can find the full tutorial below. The open access paper detailing these methods is available [here](https://doi.org/10.1080/10705511.2020.1745644) [@vanlissaTeacherCornerEvaluating2020].

If you need to install R before completing this tutorial,
I recommend following [this tutorial](https://cjvanlissa.github.io/worcs/articles/setup.html).

```{r, eval=TRUE, echo= FALSE}
blogdown::shortcode("youtube", "o750l6UUZeE")
```

## Background

A popular theory in psychology to explain social behavior is the Theory of Reasoned Action [TORA, @ajzen1980understanding]. TORA states that behavior is predicted by behavioral intention, which is in turn predicted by the attitude toward the behavior and the subjective norm about the behavior.

## Data

We examine a synthetic data set based on results found by @reineckeApplicationTheoryPlanned1996. The behavior under investigation is condom use by 16-24 year adolescents. The dependent variable ‘condom use’ is measured on a 5-point Likert-type scale (How often do you...), and the behavioral intention on a similar 5-point scale (In general do you intend to...). There are three attitude items about condom use (e.g., using a condom is awkward), three normative items (e.g., I think most of my friends would use...), all measured on a 5-point Likert-type scale.

```{r, message = FALSE}
library(foreign)
df <- read.spss("toradata.sav", to.data.frame = TRUE, use.value.labels = FALSE)
```

## Specifying the model

We have 3 indicators for attitude and norm, and observed variables for intention and behavior. We construct a measurement model for the latent variables attitude and norm.

Then, we specify a model based on TORA, with attitude and norms both predicting behavorial intention and behavior.

```{r, message = FALSE}
library(lavaan)
# Model syntax
model <- 'attit =~ attit1 + attit2 + attit3
          norm =~ norm1 + norm2 + norm3

          intent ~ attit + norm 
          behavior ~ intent + attit + norm'
# Estimate the model
fit <- sem(model, data = df)
# Examine model fit
fitmeasures(fit, fit.measures = c("npar", "chisq", "df", "rmsea", "cfi", "tli", "srmr"))
```

## Examine model results

Let's plot the model results, so we get an idea of the parameter values.

```{r}
library(tidySEM)
lo <- get_layout("",       "",       "",       "behavior", "",      "",      "",
                 "",       "",       "",       "intent",   "",      "",      "", 
                 "",       "attit",  "",       "",         "",      "norm",  "", 
                 "attit1", "attit2", "attit3", "",         "norm1", "norm2", "norm3",
                 rows = 4)
graph_sem(fit, layout = lo, angle = 179)
```

Furthermore, let's see which parameters are available for `bain` to specify hypotheses about:

```{r}
library(bain)
get_estimates(fit, standardize = TRUE)
```

## Specifying informative hypotheses

The TORA model precludes direct paths between attitude and norm and actual
behavior; the effect should be mediated totally by the behavioral intention.
Let's specify two competing hypotheses: One representing the "strict" version
of TORA, and one that also includes direct effects from attitudes and norms
on behavior. We can refer to the `get_estimates()` parameters by name.

```{r}
bain(fit, "(behavior~intent, intent~attit, intent~norm) > 0 &
           (behavior~attit, behavior~norm) = 0;
           (behavior~intent, intent~attit, intent~norm) > 0 &
           (behavior~attit, behavior~norm) > 0")
```

## Interpretation

The Bayes factors in the column labeled `BF.u` indicate the ratio of support in favor of the hypothesis, relative to any ordering of parameter values (the unconstrained hypothesis, $H_u$). Note that both informative hypotheses, $H_1$ and $H_2$, receive more support than the unconstrained hypothesis.

The Bayes factors in the column labeled `BF.c` indicate the ratio of support in favor of the hypothesis, relative to evidence not in favor of the hypothesis (the complement of the hypothesis, $H_c$). Note that both informative hypotheses receive more support than their respective complements; particularly for $H_2$ the support is overwhelming.

Perhaps most interestingly, we can examine the column labeled `PMPb`, which contains posterior model probabilities. We can use these probabilities to determine which of a set of hypotheses receives the most support. The unconstrained hypothesis is included in the set as a fail-safe: In case all informative hypotheses are a poor fit for the data, the uninformed hypothesis will receive substantial support. In this case however, we see that $H_2$ is strongly preferred over $H_1$ and $H_u$. Thus, the data cast doubt on the tenet of TORA that there are no direct effects of attitudes and norms on behavior. 