---
title: "Using WORCS to create reproducible open science projects"
author: "Caspar van Lissa"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
categories: ["R"]
tags: ["R", "Open Science", "Reproducibility"]
bibliography: [bibliography.bib]
link-citations: true
csl: apa.csl
---

This blog post provides a starting point for learning about
*WORCS: A Workflow for Open Reproducible Code in Science*. 
Below, you will find all relevant information to get you started with writing reproducible open science papers.

## Presentation

A conference presentation about WORCS is available [here on YouTube](https://www.youtube.com/watch?v=ysOxHYUWdFY). It very briefly explains the concepts behind a workflow for open science and the scope of the workflow and the accompanying R-package. The preprint paper detailing these methods is available [here]() [@vanlissaWORCSWorkflowOpen2020].

If you need to install R before completing this tutorial,
I recommend following [this tutorial](https://cjvanlissa.github.io/worcs/articles/setup.html).

```{r, eval=TRUE, echo= FALSE}
blogdown::shortcode("youtube", "ysOxHYUWdFY")
```
  

## Tutorial

In the span of a single cup of coffee, this tutorial demonstrates how to set up a new RStudio project using a `worcs` project template, and demonstrates all important features.

```{r, eval=TRUE, echo= FALSE}
blogdown::shortcode("youtube", "uzjpN_yFeUU")
```
  
## The "worcshop"

My coauthors and I give invited lectures and workshops on `worcs`.
All lecture slides and other course materials are available [here](http://cjvanlissa.github.io/worcshop).

If you want to host a worcshop at your institution, please [contact dr. Caspar J. van Lissa](mailto:c.j.vanlissa@uu.nl).

## More information

WORCS is a conceptual workflow that can be used to guide open science projects in any computational environment. The `worcs` package implements all steps in the workflow in R / RStudio, effectively offering a 'one click solution' for R-users. The package has a [website](https://cjvanlissa.github.io/worcs/), which bundles all relevant information, including the following vignettes to help users get started:

* [Setting up your computer for WORCS](https://cjvanlissa.github.io/worcs/articles/setup.html)
* [The WORCS workflow](https://cjvanlissa.github.io/worcs/articles/workflow.html)
